For submission to the Journal of Integrated Pest Management

### 2021-03-29 ofv_fife_2021.Rmd
Added images and updated files

### Abstract
Several flat mite species, all from the genus *Brevipalpus* Donnadieu (Trombidiformes: Tenuipalpidae), are the only known vectors of dichorhaviruses. The *B. californicus* species group exclusively can transmit Orchid fleck dichorhavirus (OFV) in a persistent propagative manner. OFV is the type species for the genus *Dichorhavirus* and infects more than fifty plant species belonging to the family Orchidaceae, Asparagaceae (Nolinoidaea), and Rutaceae (*Citrus*). During June 2020, chlorotic ringspot symptoms on Giant Lilyturf (*Liriope* spp., cv. ‘Gigantea’) were observed in a landscape in Leon County, Florida. Later in the year, the presence of OFV was confirmed using OFV specific conventional reverse transcription polymerase chain assay (RT-PCR) assay and Sanger sequencing. RT-PCR amplicons had a 98\% identity with the known OFV sequences available in the Genbank. The identification was also confirmed with the quantitative RT-PCR (RT-qPCR). Additional samples were taken from other Nolinoidaea, including *Liriope muscari*, *Ophiopogon japonicus*, *O. intermedius* and *Aspidistra elatior* Blume (Asparagaceae: Nolinoidaea) in Leon and Alachua counties. Identification of partial genome sequence confirmed the presence of both the orchid strains (OFV-Orc1 and OFV-Orc2) in Florida. Three mite species were recovered from OFV-infected plants: *Brevipalpus californicus* sensu lato, *B. obovatus* Donnadieu and *B. confusus* Banks. One of these species is presumably responsible for OFV transmission. Florida has various mite species of *Brevipalpus* and a diverse array of susceptible native and introduced plant species in the landscape. We suggest that OFV already has a wide distribution in Florida which will continue to spread if unchecked, representing a potential threat for *Liriope* spp., *Ophiopogon* spp. and *Aspidistra elatior* which are commonly used in landscaping in Florida . In this study we are reporting three new hosts from the family Asperagaceae. To know the extended host range of OFVs, a survey in Florida's citrus growing regions is essential, emphasizing plants within the families Rutaceae and Asparagaceae.

@@

\pagebreak
*Orchid fleck virus* (OFV) is the type member for the genus *Dichorhavirus*, family Rhabdoviridae; a bacilliform, nuclear rhabdovirus composed of two segments of single-stranded, negative-sense RNA which infects plants [@Amarasinghe2019; @Dietzgen2014; @Walker2018]. Other members of this genus are: *Citrus chlorotic spot virus*, *Citrus leprosis virus* N, *Clerodendrum chlorotic spot virus* and *Coffee ringspot virus* [@Dietzgen2018a]. Flat mites from the genus *Brevipalpus* Donnadieu (Trombidiformes: Tenuipalpidae) are the only known vector for dichorhaviruses [@Maeda1998], and mites in the *Brevipalpus californicus* (Banks) group are the only known to transmit OFV in a persistent propagative manner [@Kondo2003].

@@

During June 2020, chlorotic ringspot symptoms were observed on Giant Lilyturf *Liriope* spp., cv. 'Gigantea' in a landscape of Leon County, Florida (`r citef("fig_1")`). *Liriope* belong to a group of plants in the family Asparagaceae, subfamily Nolinoidaea, which includes a diverse array of various monocotyledonous lilliod plants which are native to southeastern Asia [@Chase2009; @Meng2021]. *Liriope* and the closely related *Ophiopogon* Ker Gawler (Asparagaceae: Nolinoidaea) are considered the most important ground cover sold by the nursery industry in southeastern US [@Mcharo2003].

@@

Further surveys of plants belonging to the subfamily Nolinoidaea in Florida have revealed more sites with symptomatic plants in Leon and Alachua counties. Mites were collected from symptomatic plants in Leon county and observed with phase contrast microscopy. We encountered both an unidentified eriophyoid mite species on the *Liriope* spp., along with flat mites on all the plant species tested (*Liriope* spp., *Ophiopogon* spp., and *A. elatior*). The flat mites were originally identified as *Brevipalpus californicus* (Banks) sensu lato by the FDACS via Differential Interference Contrast (DIC) microscopy. The *Brevipalpus* mite species complex is known to contain cryptic species [@Childers2011] that require advanced microscopy techniques, such as cryo scanning electron microscopy (Cryo-SEM) for species identification [@Skoracka2015; @Leon2010; @Beard2015]. Additional mite samples from the original OFV detection site were observed with Cryo-SEM (`r citef("fig_2")`). The determinations agreed with both prior identifications of *B. californicus* s.l. but revealed the presence of two other species *B. obovatus* and *B. confusus* (`r citef("fig_3")`).

The first report of OFV in the United States is thought to be @Ko1985, who describes nuclear inclusions caused by an undescribed bacilliform rhabdovirus in *Brassia* orchids.  The significance of this report is their reference to spoke-wheel configurations of the viral particles [@Ko1985], a sign typically associated with OFV infection [@Chang1976]. Unfortunately, @Ko1985 made no mention of mites or further investigations of this virus. The first certain report of OFV in the US was made by @Bratsch2015, confirming the presence of OFV in *Phalaenopsis* hybrids in the US, using TEM of ultrathin sections of plant tissue as well as molecular sequence analysis and its association with *Brevipalpus* mites. The authors did not make a conclusive species identification but suggested the mite vector was within the *B. californicus* group, as referred by @Kondo2003.

@@

### Conclusion
The dichorhavirus that infects citrus in Hawaii, Mexico, Colombia, and South Africa are identical to the OFV in gene order, content, and the genome sequence. According to the International committee on Taxonomy of Viruses (ICTV) classification, OFV consist of two orchid strains (OFV-Orc1 and OFV-Orc2) and two citrus strains (OFV-Cit1 and OFV-Cit2). Both the orchid strains of OFV infects citrus [@Roy2020], but none of the citrus strains have been reported from any orchid species. Detecting OFV in Florida represents a concern for horticulturists who grow orchids, *Liriope*, *Ophiopogon*, or other susceptible Asparagaceae species which are commonly used in landscaping. Florida is also home to a plethora of native and naturalized orchid species, many of which are threatened, including cultivating *Vanilla* in southern Florida [@Chambers2019] and the famous Ghost Orchid, [*Dendrophylax lindenii* (Lindl.) Benth. ex Rolfe]. Citrus leprosis was present in Florida during the 1860’s and eradicated in the mid-1960s. In fact, @Kitajima2011 found that the Citrus Leprosis virus (CiLV) which previously affected Florida citrus was a nuclear type of citrus leprosis closely related to OFV strains [@Roy2013]. Association of a distant relative of OFV named Citrus leprosis dichorhavirus-N0 (CiLV-N0) was confirmed in relation to the leprosis disease outbreak in Florida [@Hartung2015]. The recent detection of OFV-Orc1 in South Africa [@Cook2019] in *C. sinensis* (Navel and Valencia orange) and OFV-Orc2 in Hawaii [@Velarde2021] in *C. reticulata* (mandarin) and *C. jambhiri* (rough lemon) highlights the threat of different strains of OFV; which will be a definite concern to the multi-billion dollar citrus industry. Lastly, some OFV isolates are known to be involved with citrus leprosis disease in Mexico [@Roy2015], which is a cause for concern for the citrus industry. *B. californicus* and *B. yothersi* are both known vectors of Dichorhaviruses (OFV) and Cileviruses (Citrus Leprosis) [@Knorr1968; @Kondo2003; @Beltran-Beltran2020] and *B. obovatus* is a suspected vector as well [@Childers2003]. All three mite species/complexes are present in Florida [@Childers2003; @Akyazi2017] (`r citef("fig_3")`). Therefore, it is critical to identify the vector of OFVs in Florida and monitor its spread to determine the risk this virus represents for the native plants, agriculture and the ornamental/landscaping industries of Florida and the surrounding regions.

@@

<!-- ### Table -->

<!-- |Scientific Name|Common Names|Symptoms Observed| -->
<!-- |:-----------|:--------------|:----------------| -->
<!-- |*Liriope muscari* Bailey| Lilyturf, Orchardgrass, Monkeygrass| Ringspots, Chlorotic Flecking, Necrotic Lesions| -->
<!-- |*Liriope gigantea*\* Hume| Giant Lilyturf| Ringspots, Chlorotic Flecking, Necrotic Lesions| -->
<!-- |*Ophiopogon japonicus* Ker Gawl.| Dwarf Lilyturf, Mondo Grass, Snake's beard| Ringspots, Chlorotic Flecking, Necrotic Lesions| -->
<!-- |*Ophiopogon intermedius*\*\* Don| Aztec Grass, 'Argenteomarginatus' | Ringspots, Chlorotic Flecking, Necrotic Lesions| -->
<!-- |*Aspidistra elatior* Blume| Cast Iron Plant, Bar-room Plant | Chlorosis, Necrotic Lesions| -->
<!-- `r tab_nums(name = "tab_1", caption = "List of plants with symptoms of Orchid fleck dichorhavirus found in northern Florida. * *L. gigantea* have been traditionally classified as seperate from *L. muscari* by @Broussard2007 and @Fantz2015, although this distinction has been challenged by @Wang2014 and @Masiero2020. * * *O. intermedius* is sometimes misclassified as *Liriope muscari* ‘Variegated Evergreen Giant’ [@Fantz2008b, @Fantz2009] or ‘Grandiflora White’ [@Fantz2009].")` -->

>>>>

### Abstract
Several flat mite species, all from the genus *Brevipalpus* Donnadieu (Trombidiformes: Tenuipalpidae), are the only known vectors of dichorhaviruses. The *B. californicus* species group exclusively can transmit Orchid fleck dichorhavirus (OFV) in a persistent propagative manner. OFV is the type species for the genus *Dichorhavirus* and infects more than fifty plant species belonging to the family Orchidaceae, Asparagaceae (Nolinoidaea), and Rutaceae (*Citrus*). During June 2020, chlorotic ringspot symptoms on Giant Lilyturf (*Liriope* spp., cv. ‘Gigantea’, Asparagaceae: Nolinoidaea) were observed in a landscape in Leon County, Florida. Similar symptoms were observed on Giant Lilyturf (*Liriope* spp., cv. ‘Gigantea’) and Aztec Grass (*Ophiopogon intermedius* cv. 'Argenteomarginatus' D. Don, Asparagaceae: Nolinoidaea) in Alachua county, Florida. In both cases, the presence of OFV was confirmed using OFV specific conventional reverse transcription polymerase chain assay (RT-PCR) assay and Sanger sequencing. RT-PCR amplicons had a 98\% identity with the known OFV sequences available in the Genbank. The identification was also confirmed with the quantitative RT-PCR (RT-qPCR). Additional samples were taken from other Nolinoidaea, including *L. muscari*, *O. japonicus*, *O. intermedius* and *Aspidistra elatior* Blume (Asparagaceae: Nolinoidaea) in Leon and Alachua counties (`r citet("tab_1")`). Identification of partial genome sequence confirmed the presence of both the orchid strains (OFV-Orc1 and OFV-Orc2) in Florida. Three mite species were recovered from OFV-infected plants: *Brevipalpus californicus* sensu lato, *B. obovatus* Donnadieu and *B. confusus* Banks. One of these species is presumably responsible for OFV transmission. Florida has various mite species of *Brevipalpus* and a diverse array of susceptible native and introduced plant species in the landscape. We suggest that OFV already has a wide distribution in Florida which will continue to spread if unchecked, representing a potential threat for *Liriope* spp., *Ophiopogon* spp. and *Aspidistra elatior* which are commonly used in landscaping in the southeastern US. In this study we are reporting three new hosts from the family Asperagaceae. To know the extended host range of OFVs, a survey in Florida's citrus growing regions is essential, emphasizing plants within the families Rutaceae and Asparagaceae.

@@

\pagebreak
*Orchid fleck dichorhavirus* (OFV) is the type member for the genus *Dichorhavirus*, family Rhabdoviridae; a bacilliform, nuclear rhabdovirus composed of two segments of single-stranded, negative-sense RNA which infects plants [@Amarasinghe2019; @Dietzgen2014; @Walker2018]. Other members of this genus are: *Citrus chlorotic spot virus*, *Citrus leprosis virus* N, *Clerodendrum chlorotic spot virus* and *Coffee ringspot virus* [@Dietzgen2018a]. Flat mites from the genus *Brevipalpus* Donnadieu (Trombidiformes: Tenuipalpidae) are the only known vector for dichorhaviruses [@Maeda1998], and mites in the *Brevipalpus californicus* (Banks) group are the only known to transmit OFV in a persistent propagative manner [@Kondo2003].

@@

During June 2020, chlorotic ringspot symptoms were observed on Giant Lilyturf *Liriope* spp., cv. 'Gigantea' in a landscape of Leon County, Florida (`r citef("fig_1")`). *Liriope* belong to a group of plants in the family Asparagaceae, subfamily Nolinoidaea, which includes a diverse array of various monocotyledonous liliod plants which are native to southeastern Asia [@Chase2009; @Meng2021]. *Liriope* and the closely related *Ophiopogon* Ker Gawler (Asparagaceae: Nolinoidaea) are considered the most important ground cover sold by the nursery industry in southeastern US [@Mcharo2003].

@@

Further surveys of plants belonging to the subfamily Nolinoidaea in Florida have revealed more sites with symptomatic plants in Leon and Alachua counties. Mites were collected from symptomatic plants in Leon county and observed with phase contrast microscopy. We encountered both an unidentified eriophyoid mite species on the *Liriope* spp., along with flat mites on all the plant species tested (*Liriope* spp., *Ophiopogon* spp., and *A. elatior*). The flat mites were originally identified as *Brevipalpus californicus* (Banks) sensu lato by the first author and later confirmed by the FDACS via Differential Interference Contrast (DIC) microscopy. The *Brevipalpus* mite species complex is known to contain cryptic species [@Childers2011] that require advanced microscopy techniques, such as cryo scanning electron microscopy (Cryo-SEM) for species identification [@Skoracka2015; @Leon2010; @Beard2015]. Additional mite samples from the original OFV detection site were observed with Cryo-SEM (`r citef("fig_2")`). The determinations agreed with both prior identifications of *B. californicus* s.l. but revealed the presence of two other species *B. obovatus* and *B. confusus* (`r citef("fig_3")`).

The first report of OFV in the United States is thought to be @Ko1985, who describes nuclear inclusions caused by an undescribed bacilliform rhabdovirus in *Brassia* orchids.  The significance of this report is their reference to spoke-wheel configurations of the viral particles [@Ko1985], a sign typically associated with OFV infection [@Chang1976]. Unfortunately, @Ko1985 made no mention of mites or further investigations of this virus. The first certain report of OFV was from Hawaii in 2001 [@citation], while the first report from the continental US was by @Bratsch2015. In that publication, the authors confirmed the presence of OFV in *Phalaenopsis* hybrids in the US, using TEM of ultrathin sections of plant tissue as well as molecular sequence analysis and its association with *Brevipalpus* mites [@Bratsch2015]. The authors did not make a conclusive species identification but suggested the mite vector was within the *B. californicus* group, as referred by @Kondo2003.

@@

### Conclusion
The dichorhavirus that infects citrus in Hawaii, Mexico, Colombia, and South Africa are identical to the OFV in gene order, content, and the genome sequence. According to the International committee on Taxonomy of Viruses (ICTV) classification, OFV consist of two orchid strains (OFV-Orc1 and OFV-Orc2) and two citrus strains (OFV-Cit1 and OFV-Cit2). Both the orchid strains of OFV infects citrus [@Roy2020], but none of the citrus strains have been reported from any orchid species. Detecting OFV in Florida represents a concern for horticulturists who grow orchids, *Liriope*, *Ophiopogon*, or other susceptible Asparagaceae species which are commonly used in landscaping. Florida is also home to a plethora of native and naturalized orchid species, many of which are threatened, including cultivating *Vanilla* in southern Florida [@Chambers2019] and the famous Ghost Orchid, [*Dendrophylax lindenii* (Lindl.) Benth. ex Rolfe]. Citrus leprosis was present in Florida during the 1860’s and eradicated in the mid-1960s. In fact, @Kitajima2011 found that the Citrus Leprosis virus (CiLV) which previously affected Florida citrus was a nuclear type of citrus leprosis closely related to OFV strains [@Roy2013]. Association of a distant relative of OFV named Citrus leprosis dichorhavirus-N0 (CiLV-N0) was confirmed in relation to the leprosis disease outbreak in Florida [@Hartung2015]. The recent detection of OFV-Orc1 in South Africa [@Cook2019] in *C. sinensis* (Navel and Valencia orange) and OFV-Orc2 in Hawaii [@Velarde2021] in *C. reticulata* (mandarin) and *C. jambhiri* (rough lemon) associated with leprosis-like symptoms highlights the threat of different strains of OFV on citrus; which will be a definite concern to the multi-billion dollar citrus industry. Another unsettling association between *citrus* and OFV is the similarity of Mexican CiLV-N to OFV, with a genome organization identical to OFV and high phylogenetic similarity as well [@Roy2015]. *B. californicus* and *B. yothersi* are both known vectors of Dichorhaviruses (OFV) and Cileviruses (Citrus Leprosis) [@Knorr1968; @Kondo2003; @Beltran-Beltran2020] and *B. obovatus* is a suspected vector as well [@Childers2003]. All three mite species/complexes are present in Florida [@Childers2003; @Akyazi2017] (`r citef("fig_3")`). Therefore, it is critical to identify the vector of OFVs in Florida and monitor its spread to determine the risk this virus represents for the native plants, agriculture and the ornamental/landscaping industries of Florida and the surrounding regions.

@@

### Table

|Scientific Name|Common Names|Symptoms Observed|
|:-----------|:--------------|:----------------|
|*Liriope muscari* Bailey| Lilyturf, Orchardgrass, Monkeygrass| Ringspots, Chlorotic Flecking, Necrotic Lesions|
|*Liriope gigantea*\* Hume| Giant Lilyturf| Ringspots, Chlorotic Flecking, Necrotic Lesions|
|*Ophiopogon japonicus* Ker Gawl.| Dwarf Lilyturf, Mondo Grass, Snake's beard| Ringspots, Chlorotic Flecking, Necrotic Lesions|
|*Ophiopogon intermedius*\*\* Don| Aztec Grass, 'Argenteomarginatus' | Ringspots, Chlorotic Flecking, Necrotic Lesions|
|*Aspidistra elatior* Blume| Cast Iron Plant, Bar-room Plant | Chlorosis, Chlorotic Flecking, Necrotic Lesions|
`r tab_nums(name = "tab_1", caption = "List of plants with symptoms of Orchid fleck dichorhavirus found in northern Florida. * *L. gigantea* have been traditionally classified as seperate from *L. muscari* by @Broussard2007 and @Fantz2015, although this distinction has been challenged by @Wang2014 and @Masiero2020. * * *O. intermedius* is sometimes misclassified as *Liriope muscari* ‘Variegated Evergreen Giant’ [@Fantz2008b, @Fantz2009] or ‘Grandiflora White’ [@Fantz2009].")`
